---
title: "BBL Summary Statistics"
author: "Maximilian Wisnefski"
date: '2023-06-02'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(directlabels)
library(geosphere)
library(ggpubr)
library(grid)
library(scales)
library(sf)
library(sfheaders)
library(rnaturalearth)
#devtools::install_github("ropensci/rnaturalearthhires") #use if need to install pkg
library(rnaturalearthhires)
```

code author: M. Wisnefski
code editor/reviewer: S. Supp

## Introduction and read in the Bird Banding Laboratory datasets

In this code notebook, I am using the filtered BBL data to learn about the movement of individual banded birds for the 15 species of interest. I'm also making some basic visuals describing the data.

### Data filtering
In the below sections, we will filter for: 

* the correct species
* rows that represent banding events
* months that range 1:12
* individuals captured 2 or more times (recaptures, using original band)
* individuals captured in 2 or more locations (to help identify movement)

Here, we read in the data and filter out non-target species and subspecies.

```{r}
#this data has been filtered using lats/lons of ERC shapefile
#BBL_all <- readRDS("data_2023/BBL_data/BBL_in_range_2022.rds")

#this data has not been filtered
BBL_all <- readRDS("data_2023/BBL_data/BBL2022.rds")

#check the species names
BBL_all$species_name <-tolower(BBL_all$species_name)
unique(BBL_all$species_scientific_name)

#remove western subspecies and an incorrect species
BBL_all <- BBL_all %>%
  filter(!species_scientific_name %in% 
           c("Junco hyemalis oreganus", "Setophaga coronata auduboni", "Setophaga petechia"))
```

In the next step, we create a subset of the data that only includes banding events, and count the number of rows (this tells us how many individual bands are in the dataset).
```{r}
#filtering to just banding events (B), not encounters
#    since encounters all overlap with bandings
#rows with event type B also have consistent names for the different species
just_banding <- BBL_all[BBL_all$event_type == 'B',]

# check that all the species appear to be present and create a vector of names
birds <- unique(just_banding$species_scientific_name)

# summarize the counts in a table for the whole dataset (before filtering to banding events only)
bird_count <- BBL_all %>%
  group_by(species_scientific_name) %>%
  summarize(count = n())

# summarize the counts in a table for the banding only dataset
banding_count <- just_banding %>%
  group_by(species_scientific_name) %>%
  summarize(count = n())

banding_count
```

Using the BBL_all data, remove months that appear to be typos or represent something else (i.e., >12)
```{r}
#Some months are >12. We will filter that data out as it could be invalid and we aren't sure what it represents. This represents only a small amount of the observations (n=890)
table(BBL_all$event_month)

# filter to keep only months in the range 1:12
BBL_all <- BBL_all %>%
  filter(event_month %in% seq(1:12))
```

Now, what we need to know is which of the banded individuals were captured more than one time, and we can filter the dataset to only include "recaptured" individuals.
```{r}
#grouping by original band and species for all years (BBL_unique_1960)
recaptures <- BBL_all %>% 
  group_by(species_scientific_name, original_band) %>% 
  summarise(recaptures=n(), .groups = 'drop')

# get the five number summary
summary(recaptures$recaptures)
```

Keep only the individuals that were captured more than once.
```{r}
recaptures <- recaptures %>%
  filter(recaptures >= 2)
```
This filter step took our original number of bands from 4,512,246 to just 44,610 records.


Keep only the individually banded birds that were recaptured 2 or more times.
```{r}
BBL_recaptures <- BBL_all %>%
  filter(original_band %in% recaptures$original_band)
```
This filter step reduced the total number of observations from 4,565,131 to 97,495.

```{r}
#grouping previous df by unique lons and lats, as well as original id
#trying to see if any of these birds are being recaptured at different locations
unique_loc <- BBL_recaptures %>% 
  group_by(original_band, lon, lat) %>% 
  summarise(count=n(), .groups = 'drop')

#this should represent the bands that have more than one unique lat-lon. 
# sum lets us know if these are different for both. Values == 1 mean same lat and lon. Values with .5 in them mean the lat or the lon could have been the same, but the other was different.
unique_locs <- BBL_recaptures %>%
  group_by(original_band) %>%
  summarize(lats = n_distinct(lat), lons = n_distinct(lon), sum = sum(lats+lons)/2)

table(unique_locs$sum)

# now let's keep just the ones where lats > 1 and/or lons > 1
unique_locs <- unique_locs %>%
  filter(lats > 1 | lons >1)

# last, let's filter the recaptures dataset to include the individuals captured in 2 or more locations
BBL_recaptures <- BBL_recaptures %>%
  filter(original_band %in% unique_locs$original_band)
```

The filter above shows that there are, in fact, some recaptures that happen at different locations than the original, but many of these recaptures are still really close to each other.

For our final filtered data set, let's look one more time at the spread of data across the species and the timeframe (by year).
```{r}
table(BBL_recaptures$species_scientific_name)
table(BBL_recaptures$event_year)

# for table in manuscript, describing the BBL dataset used for plotting and summaries
BBL_recaptures %>%
  group_by(species_scientific_name) %>%
  summarize(indiv_count = n_distinct(original_band), 
            num_rows = n())

```


#-----------------------------------------------
### Analyze movement

This loop creates a new dataframe that collects these measures: 
* total distance (m) moved between subsequent recaptures (note that recaptures may be close or very far in time)
* bearing from point1 to point 2 (measured in degrees, proxy for n-s to e-w heading)
* East-West distance (m)
* North-South distance (m)
* whether the movement crossed 103 degrees longitude, which we're using as the division between east-west flyways

```{r}
# sort the dataset first to ensure dates are in consecutive order
BBL_recaptures <- BBL_recaptures %>%
  arrange(species_scientific_name, original_band, event_date)

# create empty data frame to store results
results <- data.frame(
  species_scientific_name = character(), orignal_band = character(), 
  date_from = character(), date_to = character(), 
  distance_m = numeric(), bearing = numeric(),
  north_south_m = numeric(), east_west_m = numeric(), 
  crossed_103_lon = logical(),
  stringsAsFactors=FALSE
)

#loop through the data by original band
for (id in unique(BBL_recaptures$original_band)) {
  # filter data for the original band (id code)
  id_data <- BBL_recaptures %>% filter(original_band == id)
  spname <- id_data$species_scientific_name[1]
  
  # loop through consecutive rows
  for (i in 1:(nrow(id_data) - 1)) {
    #coordinates of the two consecutive points
    lat1 <- id_data$lat[i]
    lat2 <- id_data$lat[i+1]
    lon1 <- id_data$lon[i]
    lon2 <- id_data$lon[i+1]
    
    #calculate great-circle distance
    point1 <- c(lon1, lat1)
    point2 <- c(lon2, lat2)
    distance <- distHaversine(point1, point2)
    
    # calculate bearing, using geosphere package
    bearing_deg <- bearing(point1, point2)
    
    #FIXME: double check that these calculations are accurate and up to date.
    #calculate North-South movement in meters
    lat_diff <- (lat2 - lat1)*111320 #1 degree latitude ~= 111.32 km
    # calculate East-West movement in meters
    lon_diff <- (lon2 - lon1)*111320*cos(lat1*pi/180) #adjust for Earth curvature
    
    # #check if the movement crossed 103 degrees longitude
    crossed_103 <- (lon1 < -103 & lon2 > -103) | (lon1 > -103 & lon2 < -103)

    # add to results
    results <- rbind(results, data.frame(
      species_scientific_name = spname, original_band = id, 
      date_from = id_data$event_date[i], 
      date_to = id_data$event_date[i+1], 
      distance_m = distance, bearing = bearing_deg,
      north_south_m = lat_diff, east_west_m = lon_diff,
      crossed_103_lon = crossed_103
    ))
  }
}

head(results)
```

there are some zero entries above because of a few individuals that had recapture events in locations very close together, such that some dates were the exact same location, and others were just slightly different. Since we want to say something about movement, let's set a lower threshold for filtering. Let's set it at 5 m, which should remove most of the really small movements that could be within the standard of error for a typical gps unit anyway.

Below we create a table with summaries for distance and bearing. Report median and iqr in paper, because the distributions are not normal.
```{r}
results <- results %>%
  filter(distance_m > 5)

species_summaries <- results%>%
  group_by(species_scientific_name) %>%
  summarize(mean(distance_m), median(distance_m), sd(distance_m), IQR(distance_m),
            mean(bearing), median(bearing), sd(bearing), IQR(bearing),
            median(abs(east_west_m)), mean(abs(east_west_m)), 
            sd(abs(east_west_m)), IQR(abs(east_west_m))
            )

species_summaries
```

count how often the observed locations cross 103 degrees longitude (indication of crossing flyways)
```{r}
table(results$species_scientific_name, results$crossed_103_lon)
```


#-----------------------------------------------
### Data visualization

Visualize the species counts, ordered from most to least observed
```{r}
plotyrcounts <- ggplot(bird_count, aes(x=reorder(species_scientific_name, -count), y=count)) +
  geom_bar(stat="identity") + 
  labs(title = "Number of individuals banded or encountered (1960-2022)",
       x = "") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        plot.title = element_text(hjust = 0.5))
```

Visualize the distribution of the species observations by years.
```{r}
# group data by species and year
year_count <- BBL_all %>% 
    group_by(species_scientific_name, event_year) %>% 
    summarise(count=n(),.groups = 'drop')

#make the line plot
plotyrs <- ggplot(year_count, 
                  aes(x = event_year, y = count, 
                      group = species_scientific_name, 
                      color = species_scientific_name)) + 
  geom_line() + 
  scale_colour_discrete(guide = 'none') +
  scale_x_discrete(expand=c(0, 1)) +
  geom_dl(aes(label = species_scientific_name), 
          method = list(dl.trans(x = x - 0.2), "first.points", cex = 0.6)) +
  theme_bw() + 
  labs(title = "Number of records 1960-2022", x = "year") +
  xlim(1950, 2021)  
```

OK let's make a nifty panel gird of the two previous plots, and save it.
```{r}
ggarrange(plotyrcounts, plotyrs, ncol=1)

ggsave("results/BBL-data-years.png", height=15, width=7)
```

Visualize the number of times individuals of each species were captured, using a histogram. Most were captured only 1-2 times. 
```{r}
# because of the distribution, we'll use a log10 y-scale to show and compare.
ggplot(recaptures, aes(x=recaptures)) +
  geom_histogram(binwidth=1) +
  scale_y_log10() +
  theme_bw() +
  facet_wrap(~species_scientific_name)
```
What we can see in the above plot is that for most species, most banded individuals are captures fewer than 10 times, with the majority only being captured once.
Mean = 1.012
Median = 1
Range = 1-28


Visualize the number of times birds changed locations
```{r}
ggplot(unique_locs, aes(x=sum)) + 
  geom_histogram(binwidth=1) + 
  scale_y_log10() +
  labs(x="number of unique locations") +
  theme_bw()
```


Visualze the months and the years that individuals were captured in the final filtered dataset
```{r}
ggplot(BBL_recaptures, aes(x=event_month)) +
  geom_histogram(binwidth=1) +
  scale_x_continuous(breaks= pretty_breaks()) +
  labs(x="Month observed") +
  theme_bw() +
  facet_wrap(~species_scientific_name)

ggplot(BBL_recaptures, aes(x=event_year)) +
  geom_histogram(binwidth=1) +
  scale_x_continuous(breaks= pretty_breaks()) +
  labs(x="Month observed") +
  theme_bw() +
  facet_wrap(~species_scientific_name)
```


### Creating maps to visualize individual movement and recapture data

Let's get the underlying shapefiles for country boundaries
```{r}
#making map of range that I am interested in
state_prov <- rnaturalearth::ne_states(c("united states of america", "canada", "mexico", "honduras", "el salvador", "guatemala", "costa rica"), returnclass = "sf")
#state_prov <- st_crop(state_prov, xmin = -115, xmax = -50, ymin = 5, ymax = 50)
state_prov <- st_crop(state_prov, xmin = -140, xmax = -50, ymin = 5, ymax = 60)
```

This code loops through all species for 1960-2022 to make lines between their recapture locations. This provides a very coarse look at the movement directions individuals may take within their lifetimes.
```{r}
#mapping the movements of recaptured individuals 
i = 1 
birds <- unique(just_banding$species_scientific_name)
for (b in birds){
  #these next two lines are so I can also include common name on map
  name = birds[i]
  i = i + 1

  df <- BBL_recaptures[BBL_recaptures$species_scientific_name == b, ]
  df <- df[df$lat != 0.0, ] 
  
  #setting color to original_band because each original band represents an individual bird
  print(ggplot() + 
    geom_sf(data=state_prov, colour="grey50", fill="white") + 
    geom_path(data = df, aes(x = lon, y = lat, group=original_band), 
              col = "olivedrab",
              linewidth = 0.25, lineend = "round", alpha=0.4) +
    labs(x = "", y = "", title = b) +
    theme_minimal() +
    theme(legend.position = "none", plot.title = element_text(hjust = 0.5), 
          plot.subtitle = element_text(hjust = 0.5)) +
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank() 
          ) +
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
    )
}

```

Show the ew vs ns movement in a scatterplot
```{r}
ggplot(results, aes(x=(abs(east_west_m/1000)), y=(abs(north_south_m/1000)))) + 
  geom_point(aes(col = ifelse(abs(north_south_m/1000)>abs(east_west_m/1000), 
                              "n-s", "e-w")), alpha=0.4) + 
  geom_abline(intercept=0, slope=1, linetype='dashed') +
  scale_color_manual(values = c("n-s" = "skyblue4", "e-w" = "orange3")) +
  facet_wrap(~species_scientific_name) + 
  xlab("east-west movement (km)") + 
  ylab("north-south movement (km)") +
  theme_bw() + 
  theme(legend.position = "none")
```

Show the bearings (degree) in a histogram
```{r}
ggplot(results, aes(x=bearing)) + 
  geom_histogram(binwidth=20) + 
  facet_wrap(~species_scientific_name) +
  xlim(-180, 180) + 
  geom_vline(xintercept = c(0, -90, 90), linetype="dashed") +
  theme_bw()

ggsave("results/BBL_bearing.png")
```






#--------------------------------------- NOT PART OF THE MAIN MANUSCRIPT

## Analysis only for recent data (2008-2022). This was done as part of the summer study, but we used all data (1960-2022) for the main manuscript.

### data subset for 2008-2022
Let's also make a subset of the data that represents the same years as the recent eBird occurrences, just for a more direct comparison
```{r}
#should probably also get rid of 2022 since there is not data for full year
BBL_recent <- BBL_all[BBL_all$event_year >= 2008,]
just_banding_recent <- just_banding[just_banding$event_year >= 2008, ]

#getting counts for each bird
# summarize the counts in a table for the whole dataset
recent_count <- BBL_recent %>%
  group_by(species_scientific_name) %>%
  summarize(count = n())
```

### show number of individuals for each recent year
Let's also make the plot for the recent years of data, for comparison
```{r}
ggplot(bird_count, aes(x=reorder(species_scientific_name, -count), y=count)) +
  geom_bar(stat="identity") + 
  labs(title = "Number of individuals banded or encountered (2008-2022)",
       x = "") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        plot.title = element_text(hjust = 0.5))
```

### look at how counts by species change in the recent years
Let's look at this again, but just with the most recent years, 2008-2022.
```{r}
#grouping by year and species for just more recent years
#using just_banding_recent because it has consistent names for all species, so the groupby actually works
year_count_recent <- BBL_recent %>% 
  group_by(species_scientific_name, event_year) %>% 
  summarise(count=n(),.groups = 'drop')

# make the line plot
ggplot(year_count_recent, aes(x = event_year, y = count, 
                       group = species_scientific_name, color = species_scientific_name)) + 
  geom_line() + 
  scale_colour_discrete(guide = 'none') +
  scale_x_discrete(expand=c(0, 1)) +
 # geom_dl(aes(label = species_scientific_name), method = list(dl.combine("first.points", "last.points")), cex = 0.8) +
  #geom_dl(aes(label = species_scientific_name), method = list(dl.trans(x = x + 0.1), "last.points", cex = 0.8)) +
  geom_dl(aes(label = species_scientific_name), method = list(dl.trans(x = x - 0.2), "first.points", cex = 0.6)) +
   theme_bw() + 
  labs(title = "Number of records 2008-2022",
       x = "year") +
  xlim(2005, 2021)  
```

This code focuses on data from 2008 or later for maps showing movement between locations.
```{r}
#filtering recapture data to only include 2008 and later 
BBL_recaptures_recent <- BBL_recaptures[BBL_recaptures$event_year >= 2008,]

#mapping the movements of recpatured individuals 
i = 1 
for (b in birds){
  #these next two lines are so I can also include common name on map
  name = birds[i]
  i = i + 1

  df <- BBL_recaptures_recent[BBL_recaptures_recent$species_scientific_name == b, ]
  df <- df[df$lat != 0.0, ] 
  
  #setting color to original_band because each original band represents an individual bird
  print(ggplot() + 
    geom_sf(data=state_prov, colour="grey50", fill="white") + 
    geom_path(data = df, aes(x = lon, y = lat, col=original_band), 
              linewidth = 0.5, lineend = "round") +
    labs(x = "", y = "", title = b) +
    theme_minimal() +
    theme(legend.position = "none", plot.title = element_text(hjust = 0.5), 
          plot.subtitle = element_text(hjust = 0.5)) +
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank() 
          ) +
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
    )
  #ggsave(paste0("figs/BBL_", b, ".png"), bg = "transparent", height = 5, width = 7.5)

}

```
