---
title: "seed-disperser-diversity"
author: "Sarah Supp"
date: "2025-03-03"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

##Code for Eastern redcedar avian seed disperser project (c) 2025, Supp
(PI), Wisnefski, and La Sorte
[supps\@denison.edu](mailto:supps@denison.edu){.email} Denison
University Code is under development, part of NSF Multi-Institution
Collaborative Award (2019-2025)

Birds evaluated include:

-   Cedar Waxwing (CEWA) *Bombycilla cedrorum*
-   Robin (ROBI) *Turdus migratorius*
-   Wood Thrush (WOTH) *Hylocichla mustelina*
-   Yellow-rumped Warbler (YEWA) *setophaga coronata*
-   Blue Jay (BLJA) *Cyanocitta cristata*
-   European Starling (EUST) *Sturnus vulgaris*
-   Eastern Bluebird (EABL) *Sialia sialis*
-   Northern Mockingbird (NOMO) *Mimus polyglottos*
-   Downy Woodpecker (DOWO) *Dryobates pubescens*
-   Eastern Meadowlark (EAME) *Sturnella magna*
-   White-breasted Nuthatch (WHNU) *Sitta carolinensis*
-   Purple Finch (PUFI) *Haemorhous purpureus*
-   Northern Cardinal (NOCA) *Cardinalis cardinalis*
-   Dark-eyed Junco (DAJU) *Junco hyemalis*
-   American Crow (AMCR) *Corvus brachyrhynchos*

The inputs to this code file are the results from a community-level
analysis, shared by collaborator La Sorte. This includes for each
hexagon cell (POLYFID) and year (16 winter seasons from 2008 to 2023)
the number of target species (S; range=0-15), three diversity metrics
(est; richness, shannon, simpson), and survey completeness estimates
(sc). Values were calculated using the iNEXT package for the analysis
and the same eBird data used in the population-level analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(broom)
library(dggridR)
library(ggpubr)
library(ggridges)
library(here)
library(sf)
library(tidyverse)
library(viridis)
```

## Import data

Import the iNext data for the community analysis. It is based on the
eBird occurrence information for all 15 seed dispersing species. Winter
season was defined as November through February. We used years
2008-2023.

```{r}
# diversity estimates for winter seasons, by POLYFID hexagonal cell; loads 3 dataframes
load(here("data/1-raw_eBird_data/community/iNEXT-results.RData"))

# coordinates for the icosahedron cells
load(here("data/1-raw_eBird_data/locs.RData")) 
```

Import the shapefile for the eastern redcedar range extent boundaries.
This code file imports a shapefile representing an updated range
boundary for the eastern redcedar tree, Juniperus virginia. The
shapefile was created by W. Haas, 2023 (Kent State University).

```{r}
ERC_range_sf <- st_read(here("data/shapefiles/ERC_shapefile_2023/junivirg.shp"), quiet = TRUE) %>%
  st_transform(crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```

## Process the data

merge the diversity files with the location file so that polygon
locations can be mapped

```{r}
# use merge to annotate SURVEY COMPLETENESS with the polygon coordinates (lat, long)
sc <- merge(sc, locs)
```

Import map polygons for plotting

```{r}
#Get polygons for the spatial range and make a map
north_america <- map_data("world") %>%
  filter(region %in% c("USA", "Canada", "Mexico", "Guatemala", "Belize", 
                       "El Salvador", "Honduras", "Nicaragua", "Costa Rica", 
                       "Panama"))
```

Take the sc dataframe and use dggridr package to prepare grid cell data
for mapping hexagons and their values. This code also adds columns that
label the rows as "include" or "exclude", based on the survey
completeness threshold \>= 90%; and a column that contains the SC values
when \>=90% but fills in NA for all values below the threshold. This is
to make a map that clearly shows which values will be used, and which
ones will be filtered out.

```{r}
#Prepare the grid cell data for mapping

# use the suvery completeness dataframe; and cell centers calculated
dgg <- dgconstruct(project = "FULLER", aperture = 4, topology = "HEXAGON", res = 6)
sc$cell <- dgGEO_to_SEQNUM(dgg, sc$x, sc$y)$seqnum
sc$cell_lat <- dgSEQNUM_to_GEO(dgg, sc$cell)$lat_deg 
sc$cell_lon <- dgSEQNUM_to_GEO(dgg, sc$cell)$lon_deg 

# Get the grid cell boundaries 
grid <- dggridR::dgcellstogrid(dgg, sc$cell) 

#Update the grid cells' properties to include sc values
grid <- merge(grid, sc, by.x="seqnum", by.y="cell")

# #Make adjustments so the output is more visually interesting
grid <- grid %>% mutate(threshold=ifelse(SC<0.90, "exclude", "include"))
grid <- grid %>% mutate(thres_vals=ifelse(SC<0.90, NA, round(SC, 2)))
```

Identify which cells fall within and outside of the eastern redcedar
range extent (ERC_range_sf).

-   Convert the survey completeness (sc) dataframe into an sf object.
-   Geometry for sc_sf represents the center point of the icosahedron
    cell within which the observation took place.
-   Geometry for ERC_range_sf represents the range boundaries for
    eastern red cedar tree

```{r}
#convert sc grid data into an sf object 
sc_sf <- st_as_sf(grid, 
                     coords = c("cell_lon", "cell_lat"), 
                     crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

# apply st_intersects to sc data
sc_sf_ERC = lengths(st_intersects(sc_sf, ERC_range_sf)) > 0

# use sc_sf_ERC to filter the sc data to only include observations WITHIN ERC RANGE extent
sc_ERC_in = sc_sf %>%
  filter(geometry %in% geometry[sc_sf_ERC]) %>%
  mutate(ERC="within_range")

#use sc_sf_ERC to filter the sc data that falls OUTSIDE ERC RANGE extent
sc_ERC_out = sc_sf %>%
  filter(! geometry %in% geometry[sc_sf_ERC]) %>%
  mutate(ERC="outside_range")

sc_ERC <- rbind(sc_ERC_in, sc_ERC_out)
```

merge the sc_ERC dataframe with the S and est dataframes. This will add
the grid annotations and the ERC range labels to both of the other
dataframes.

```{r}
# merge in the order shown, because then the resulting dataframe will inherit the status of being an sf object. If merged in the opposite order, the resulting dataframe will no longer be an sf object.
S_ERC <- merge(sc_ERC, S)

est_ERC <- merge(sc_ERC, est)
```

#--------------------------------------------------

# Analyze and visualize the data

## Survey Completeness

Survey completeness of the bird occurrence information was estimated
using the Hill numbers approach (Chao et al., 2014). We implement this
analysis for each winter season using the iNEXT R package (Hsieh et al.,
2024). We only included checklists in the survey completeness analysis
where all the species detected by sight or sound were recorded by the
observer (i.e. the ‘all observations reported’ field in eBird was
designated as ‘yes’) (La Sorte et al., 2024). For our community
analysis, we only examined spatial bins that were well surveyed in each
season, with a survey completeness threshold \>=90% (La Sorte et al.,
2024).

Look at the survey completeness data, by year. Here, we use the
dataframe that represents the cells falling within the ERC range

```{r}
ggplot(data=sc_ERC %>% filter(ERC=="within_range"), 
       aes(x=SC, y=as.factor(year), group=year, fill=stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "survey completeness", option = "D") +
  labs(y = "Winter Season Year",
       x = "Survey Completeness") +
  geom_vline(aes(xintercept=0.90)) +
  theme_ridges() +
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8))
```

There is no major pattern in survey completeness across time, but maybe
a slight increase in completeness in more recent years. Within the ERC
range, most cells are very well sampled.

Check the relationship of observed species richness with survey
completeness. Here, we use the dataframe that represents the cells
falling within the ERC range.

```{r}
ggplot(sc_ERC %>% filter(ERC == "within_range"), aes(S.obs, SC)) +
  geom_point(alpha=0.10) + 
  xlab("Observed species richness") +
  ylab("Survey Completeness") +
  theme_bw()
```

As species richness increases, it is more likely that high sample
completeness will be estimated. There are still a good number of low
species richness sites that also reach high survey completeness.

Let's plot survey completeness on a map to show the estimates
transparently. Note: Here we use the previously created "grid" dataframe
version of "sc" for the polygons and the fill values. We overlay a
polygon showing the ERC range extent on top of the hexagons.

```{r}
# Winter survey completeness per grid cell (Nov-Feb)
ggplot() + 
  geom_sf(data = grid, aes(fill = thres_vals), color = alpha("white", 0.4 )) +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black", linewidth=0.10) +
  geom_sf(data = ERC_range_sf, fill = NA, color ="black", linewidth=0.50) +
  #scale_fill_gradient(low="grey90", high="black", na.value="white") +
  scale_fill_gradient(low="yellow", high="red") +
  coord_sf(xlim = c(-103, -60), ylim = c(20, 55)) + #set limits without distorting boundaries
  labs(title = "Winter survey completeness per grid cell (Nov-Feb)") +
  theme_minimal() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          legend.position="bottom") +
  facet_wrap(~year, ncol=4)

ggsave(filename = here("figs/sample-coverage_map_08-23.png"), height = 11, width=8, bg="white")
```

## Seed disperser species diversity over time

The S dataframe counts the number of target species (0-15) that were
observed in each cell during a winter season (November-February). S_ERC
has already been merged with the gridded data (for dggridr) and labled
as within or outside of the ERC range.

### For composition/diversity analysis, use the survey completeness threshold \>=0.90 (90%) AND filter to cells within the ERC range extent.

Filter the S and est dataframes. Cells \<0.90 we cannot assume to be
well sampled.

```{r}
# filter observations to only include those with >=90% survey completeness and within ERC Range
S_ERC_90sc <- S_ERC %>%
  filter(SC >= 0.90 & ERC=="within_range") %>%
  arrange(POLYFID, year)

est_ERC_90sc <- est_ERC %>%
  filter(SC >= 0.90 & ERC=="within_range") %>%
  arrange(POLYFID, year)

```


# Species richness

**Questions:**

-   Spatial: Are there areas with increased richness of ERC seed
    dispersers?
    -   A hypothesis is that there should be higher richness of ERC seed
        dispersers in the southeastern united states, and in east coast
        urban areas (heat island, feeder effects, etc.)
-   Temporal: Is seed disperser richness stable through time, or is
    there evidence of overall increase?
    -   A hypothesis could be that changing climate is increasing seed
        disperser richness during the winter because of advanced
        migration and/or increased number of individuals or
        subpopulations that are not migrating southward, out of the ERC
        range
    -   A spatial component could include a shift in areas that are
        experiencing this trend, and other areas that are not.

```{r}
ggplot(data=S_ERC_90sc, aes(x=S, y=as.factor(year), group=year, fill=stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "species richness", option = "C") +
  labs(y = "Winter Season Year",
       x = "Species Richness") +
  theme_bw() +
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8))
```

Show species richness on a map, separated by year

```{r}
# Winter species richness per grid cell (Nov-Feb)
ggplot() + 
  geom_sf(data = S_ERC_90sc, aes(fill=S), color = alpha("white", 0.4 )) +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black", linewidth=0.10) +
  geom_sf(data = ERC_range_sf, fill = NA, color ="black", linewidth=0.25) +  
  scale_fill_viridis_c(option = "plasma") +
  coord_sf(xlim = c(-106, -70), ylim = c(29, 49)) + #set limits without distorting boundaries
  labs(title = "Winter species richness per grid cell (Nov-Feb)") +
  theme_minimal() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          legend.position="bottom") +
  facet_wrap(~year, ncol=4)

ggsave(filename = here("figs/S_map_08-23.png"), height = 8, width=8, bg="white")

```

### Is there a significant trend for changing species richness across cells, through time?

Calculate the slopes of change across years, within cells. Plot the
results to see if there is a trend of higher species disperser load
through time. Need to think and check with FAL for ideas to see if we
need to account for overall effort better (currently using as pres-only
data, without weighted checks)

**Note:** If fewer than half the years are present, then the linear model for that cell will be skipped.
#FIXME: Decide if we want to keep or relax this requirement.
```{r}
# calculate separate lm for all grid cells
cells <- unique(S_ERC_90sc$POLYFID)

lm_S_results <- data.frame(POLYFID=numeric(), n=integer(), slope=numeric(), 
                           lci=numeric(), uci=numeric(), pVal=numeric())

for (c in cells) {
  dat <- S_ERC_90sc %>% filter(POLYFID == c)
  #must have at least half of the years (n=8) to run a linear model
  if (nrow(dat) >= 8) {
      mod <- lm(S~as.numeric(year), data=dat)
      n = nrow(dat)
      slope = coef(mod)[[2]]
      lci = confint(mod)[2,1]
      uci = confint(mod)[2,2]
      pVal = glance(mod)$p.value[[1]]
      
      newrow <- data.frame(POLYFID=c, n, slope, lci, uci, pVal)
      lm_S_results <- rbind(lm_S_results, newrow)
  }
}

#remove the NA values
lm_S_results <- na.omit(lm_S_results)
  
# add column for significance
lm_S_results$sig <- ifelse(lm_S_results$pVal <= 0.05, "sig", "NS")

# add the cell locations back into the dataset
unique_cell_dat <- S_ERC_90sc %>%
  select(POLYFID, seqnum, cell_lon, cell_lat, geometry) %>%
  distinct()

#merge results with cell data
lm_S_results <- merge(lm_S_results, unique_cell_dat)

#transform back into sf object
lm_S_results <- st_as_sf(lm_S_results, crs = 4326)
```

Plot the results from the S linear regression

```{r}
#ggplot histogram, layered by identity for sig and NS
ggplot(lm_S_results, aes(slope)) + 
  geom_density(aes(fill=sig), alpha=0.50) +
 # geom_histogram(aes(fill=sig), binwidth=0.05, position="identity", alpha=0.5) +
  scale_fill_manual(values = c("gray40", "navyblue")) + 
  geom_vline(xintercept=0, linetype="dashed", col="gray10") +
  xlab("slope estimate") + 
  theme_bw()

ggsave(filename = here("figs/S_slope_density-plot.png"), height = 3, width=4, bg="white")
```

Most of the significant slopes were positive (increasing richness). Very
few significant negative slopes (decreasing richness).

```{r}
#average effect size, mean of the slopes
avg_effect_size <- lm_S_results %>% 
  filter(sig=="sig") %>%
  summarize(meanslope = mean(slope),
            medianslope = median(slope))

avg_effect_size
```

See if there is a signal by longitude (would indicate a E-W influence or trend)
```{r}
ggscatter(data=lm_S_results, x="cell_lon", y="slope", add="reg.line", alpha=0.25, col="sig") +
  stat_cor(label.y = 0.8, aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  stat_regline_equation(label.y = 0.7) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  scale_colour_manual(values = c("gray40", "navyblue")) + 
  theme(text = element_text(size=9), axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  #scale_x_continuous(breaks=seq(2008, 2024, 6)) +
  xlab("Cell Longitude") + ylab("S regression slope") + 
  theme_bw()

ggsave(filename = here("figs/S_slope_scatter-longitude.png"), height = 3, width=4, bg="white")

```

See if there is a signal by latitude (would indicate a N-S influence or trend)

```{r}
ggscatter(data=lm_S_results, x="cell_lat", y="slope", add="reg.line", alpha=0.25, col="sig") +
  stat_cor(label.y = 0.8, aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  stat_regline_equation(label.y = 0.7) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  scale_colour_manual(values = c("gray40", "navyblue")) + 
  theme(text = element_text(size=9), axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  #scale_x_continuous(breaks=seq(2008, 2024, 6)) +
  xlab("Cell Latitude") + ylab("S regression slope") + 
  theme_bw()

ggsave(filename = here("figs/S_slope_scatter-latitude.png"), height = 3, width=4, bg="white")

```

Plot the slope estimates against S in year 1 (2008). 
```{r}
initial_S <- S_ERC_90sc %>%
  filter(year == 2008) %>%
  select(POLYFID, S)

Sxslope <- st_join(lm_S_results, initial_S)

ggplot(Sxslope, aes(S, slope)) + 
  geom_jitter(aes(col=sig), alpha=0.20) +
  scale_colour_manual(values = c("gray40", "navyblue")) + 
  theme_bw() +
  xlab("initial richness (S) in 2008")

ggsave(filename = here("figs/slope_initial_S.png"), height = 3, width=4, bg="white")
```


Let's plot the map using the significant slopes, and all others shaded
in grey
```{r}
# make subsets based on significant vs non-significant slope pvalues
gridsig <- lm_S_results %>% filter(sig == "sig")
gridNS <- lm_S_results %>% filter(sig == "NS")


# Winter species richness change (slope), per grid cell (Nov-Feb)
ggplot() + 
  geom_sf(data = gridsig, aes(fill = slope), color = alpha("white", 0.4 )) +
  geom_sf(data = gridNS, fill="grey", color = alpha("white", 0.04)) + 
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black", linewidth=0.10) +
  geom_sf(data = ERC_range_sf, fill = NA, color ="black", linewidth=0.50) +
  scale_fill_gradient2(low="firebrick3", high="navyblue", midpoint = 0) +
  coord_sf(xlim = c(-106, -70), ylim = c(29, 49)) + #set limits without distorting boundaries
  labs(title = "Winter species richness slope per grid cell (2008-2023)") +
  theme_void() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          legend.position="bottom")

ggsave(filename = here("figs/S_slope_map.png"), height = 4, width=6, bg="white")

```
Target species richness is increasing just about everywhere where significant trends are found. Many are non-significant, but only a very few significant trends are negative (n=1).


#FIXME: Next step: what do do with the estimate dataset
