---
title: "ERC shapefile basics"
author: "Maximilian Wisnefski"
date: '2023-05-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sf)
library(mapdata)
```

# Introduction
This code file imports a shapefile representing an updated range boundary for the eastern redcedar tree, Juniperus virginia. The shapefile was created by W. Haas, 2023 (Kent State University).
The file also imports occurrence data from eBird and the Breeding Bird Laboratory. 
It uses the shapefile as a filter to identify occurrences that fall within the eastern redcedar (ERC) species range.

The file was originally authored by M. Wisnefski (Anderson Summer Scholar, 2023) and updated by S. Supp (2024-2025).


## Prepare the Eastern Redcedar range data

Import the ERC range file, updated by W. Haas, 2022
```{r}
ERC_range_sf <- st_read(here("data/shapefiles/ERC_shapefile_2023/junivirg.shp"), quiet = TRUE) %>%
  st_transform(crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```

Plot the shapefile on a map of North America.
Get the map boundaries.
```{r}
#Get polygons for the spatial range and make a map
north_america <- map_data("world") %>%
  filter(region %in% c("USA", "Canada", "Mexico"))
```

plot the range shapefile on a map to check that it's working
```{r}
ggplot() + 
  geom_sf(data = ERC_range_sf, fill = "darkolivegreen3", color = alpha("darkolivegreen")) +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black") +
  coord_sf(xlim = c(-125, -70), ylim = c(25, 50)) + #show within all north america
 # coord_sf(xlim = c(-105, -70), ylim = c(25, 50)) + #focus on eastern us
 # labs(title = "W. Haas shapefile of J. virginiana range extent") +
  theme_void() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank())

ggsave(filename = here("figs/ERC_range_map.png"), height = 4, width=5.5)
```


## Filter within range eBird data, 2008-2023

Read in the eBird data that was summarized in previous code
```{r}
eBird_2023 <- readRDS("data_2023/dat_effort_2023.rds")
```

Convert the eBird dataframe into an sf object
```{r}
#convert eBird data into an sf object 
eBird_sf <- st_as_sf(eBird_2023, 
                     coords = c("x", "y"), 
                     crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```

Find out which avian occurrences fall within the eastern redcedar range (ERC_range_sf)
Geometry for eBird_sf represents the center point of the icosahedron cell within which the observation took place.
Geometry for ERC_range_sf represents the range boundaries for eastern red cedar tree

Note: The filter step requires several minutes to run on a desktop computer.

WITHIN RANGE DATA
```{r}
# apply st_intersects to eBird data
inside_eBird = lengths(st_intersects(eBird_sf, ERC_range_sf)) > 0

# use inside_eBird to filter the eBird data to only include observations within ERC range extent
eBird_in_range = eBird_sf %>%
  filter(eBird_sf$geometry %in% eBird_sf$geometry[inside_eBird])
```

clean up the columns
```{r}
#split the geometry column back into lat col and lon col by converting sf object back to df
eBird_in_range <- sf_to_df(eBird_in_range, fill = TRUE )

#remove unnecessary columns created by above conversion
eBird_in_range = subset(eBird_in_range, select = -c(sfg_id, point_id))
```

OUTSIDE OF RANGE DATA
```{r}
# Identify which points are outside the ERC range extent
eBird_out_range = eBird_sf %>%
  filter(! eBird_sf$geometry %in% eBird_sf$geometry[inside_eBird])

#split the geometry column back into lat col and lon col by converting sf object back to df
eBird_out_range <- sf_to_df(eBird_out_range, fill = TRUE )

#remove unnecessary columns created by above conversion
eBird_out_range = subset(eBird_out_range, select = -c(sfg_id, point_id))
```

Save filtered data as a file to the main project folder
```{r}
#export as .rds
saveRDS(eBird_in_range, "data_2023/dat_effort_in_range_2023.rds")
```



## Filter within range BBL data, 2008-2023
Read in the BBL data and conduct some basic data preparation steps
```{r}
BBL <- readRDS("data_2023/BBL_data/BBL2022.rds")

#change names of lat/lon columns for convenience
colnames(BBL)[which(names(BBL) == "lat_dd")] <- "lat"
colnames(BBL)[which(names(BBL) == "lon_dd")] <- "lon"

#remove rows with NA for lat/lon columns
BBL <- BBL[!is.na(BBL$lat),]
BBL <- BBL[!is.na(BBL$lat),]

# convert the BBL dataframe into an sf object
BBL_sf <- st_as_sf(BBL, coords = c("lon", "lat"), crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```

Find out which BBL records occur within the ERC range
```{r}
#apply st_intersects to BBL data
inside <- lengths(st_intersects(BBL_sf, ERC_range_sf)) > 0

BBL_in_range <- BBL_sf %>%
  filter(BBL_sf$geometry %in% BBL_sf$geometry[inside])
```

Identify which BBL records ARE NOT within the ERC range
```{r}
BBL_out_range <- BBL_sf %>%
  filter(! BBL_sf$geometry %in% BBL_sf$geometry[inside])
```

clean up the filtered data to write back to tabular format for storing
```{r}
#  WITHIN RANGE DATA
#split the geometry column back into lat col and lon col by converting sf object back to df
BBL_in_range <- sf_to_df(BBL_in_range, fill = TRUE )

#remove unnecessary columns created by above conversion
BBL_in_range = subset(BBL_in_range, select = -c(sfg_id, point_id))

#rename columns
BBL_in_range <- rename(BBL_in_range, lon = x)
BBL_in_range <- rename(BBL_in_range, lat = y)
```

```{r}
#  OUTSIDE RANGE DATA
#split the geometry column back into lat col and lon col by converting sf object back to df
BBL_out_range <- sf_to_df(BBL_out_range, fill = TRUE )

#remove unnecessary columns created by above conversion
BBL_out_range = subset(BBL_out_range, select = -c(sfg_id, point_id))

#rename columns
BBL_out_range <- rename(BBL_out_range, lon = x)
BBL_out_range <- rename(BBL_out_range, lat = y)
```

Save filtered data as a file to the main project folder
```{r}
#export as .rds
saveRDS(BBL_in_range, "data_2023/BBL_data/BBL_in_range_2024.rds")
```

## Final plots of the eBird data and the BBL data within the ERC range

### eBird data maps 
Note: there is a lot of overplotting because the points represent cell centers, not individual observation coordinates
black = within the range, red = outside of the range
```{r}
# plot the checklists submitted within the range
ggplot() + 
  geom_sf(data = ERC_range_sf, fill = "lightgreen", color = alpha("darkgreen")) +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black") +
  geom_point(data=eBird_in_range, aes(x, y), alpha=0.25, size=0.1) +
  geom_point(data=eBird_out_range, aes(x, y), alpha=0.25, size=0.1, col="indianred") + 
  coord_sf(xlim = c(-103, -65), ylim = c(25, 50)) + #set limits without distorting boundaries
  labs(title = "eBird occurrences within J. virginiana range extent") +
  theme_minimal() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank())

ggsave(filename = "figs/eBird_and_range_map.png", height = 4, width=5.5)
```

### BBL data maps
black = within the range, red = outside of the range
```{r}
# plot the number of records submitted within the range
ggplot() + 
  geom_sf(data = ERC_range_sf, fill = "lightgreen", color = alpha("darkgreen")) +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black") +
  geom_point(data=BBL_in_range, aes(lon, lat), alpha=0.25, size=0.1) +
  geom_point(data=BBL_out_range, aes(lon, lat), alpha=0.25, size=0.1, col="indianred") +
  coord_sf(xlim = c(-103, -65), ylim = c(25, 50)) + #set limits without distorting boundaries
  labs(title = "BBL records within J. virginiana range extent") +
  theme_minimal() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank())

ggsave(filename = "figs/BBL_and_range_map.png", height = 4, width=5.5)

```
