---
title: "figures"
author: "Sarah Supp"
date: "2025-03-07"
output: html_document
---

I'm using this file to generate figures and maps to accompany the eBird + ERC manuscript

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dggridR)
library(here)
library(lubridate)
library(sf)
library(tidyverse)
library(viridis)
```

import species and geographic range data
```{r}
#import the species counts and effort data
dat_effort <- readRDS(here("data/2-intermediate_eBird_data/dat_effort_filtered.rds"))

#import the J. virginiana geographic range
ERC_range_sf <- st_read(here("data/shapefiles/ERC_shapefile_2023/junivirg.shp"), quiet = TRUE) %>%
  st_transform(crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```

generate map boundaries
```{r}
#Get polygons for the spatial range and make a map
north_america <- map_data("world") %>%
  filter(region %in% c("USA", "Canada", "Mexico"))
```

#Map an overview of winter occurrence for (1) fruit and (2) winter fruit feeder species

**fruit feeder species**
-   B. cedrorum
-   C. cardinalis
-   T. migratorius

**Winter fruit feeder species**
-   H. mustelina
-   M. polyglottus
-   S. coronata
-   S. sialis
-   S. carolinensis

**Opportunistic fruit feeder species**
-   Corvus brachyrhynchos
-   Cyanocitta cristata
-   Dryobates pubescens
-   Haemorhous purpureus
-   Junco hyemalis
-   Sturnella magna
-   Sturnus vulgaris

make a version of effort that pools across years, just to show the general maps. Need to do some general processing first here. 
```{r}
#subset to just the species that are winter fruit feeders
 wff <- dat_effort # %>%
#   filter(species %in% c("Bombycilla_cedrorum", "Cardinalis_cardinalis",
#                         "Turdus_migratorius",
#                         "Hylocichla_mustelina", "Mimus_polyglottus", 
#                         "Setophaga_coronata", "Sialia_sialis", 
#                         "Sitta_carolinensis"))

# add a column for month, to identify winter fruiting season
unique_dates <- wff %>%
  distinct(year, day) %>%
  mutate(month = month(as.Date(day - 1, origin = paste0(year, "-01-01"))))
# join back to the main dataframe
wff <- wff %>%
  left_join(unique_dates, by = c("year", "day"))

# keep only the winter months (October-March)
wff <- wff %>%
  filter(month %in% c(10, 11, 12, 1, 2, 3))

#make annual winter summary of the frequency of observation for the wff species
wff_sum <- wff %>%
  group_by(species, POLYFID, x, y) %>%
  summarize(sumcount = sum(count), sumeffort = sum(effort)) %>%
    mutate(freq= sumcount/sumeffort)

# label species by diet
wff_sum <- wff_sum %>%
  mutate(diet = ifelse(species %in% 
                         c("Bombycilla_cedrorum", "Cardinalis_cardinalis",
                           "Turdus_migratorius"),"fruit feeder", 
                       ifelse( species %in% c("Hylocichla_mustelina", 
                               "Mimus_polyglottus", "Setophaga_coronata",
                               "Sialia_sialis", "Sitta_carolinensis"), 
                       "winter fruit feeder",
                       "opportunistic")))

```


Take the wff dataframe and use dggridr package to prepare grid cell data
for mapping hexagon cells and their values. This is
to make a map that clearly shows which values will be used, and which
ones will be filtered out.

```{r}
#Prepare the grid cell data for mapping

# use the processed dataframe; and cell centers calculated
dgg <- dgconstruct(project = "FULLER", aperture = 4, topology = "HEXAGON", res = 6)
wff_sum$cell <- dgGEO_to_SEQNUM(dgg, wff_sum$x, wff_sum$y)$seqnum
wff_sum$cell_lat <- dgSEQNUM_to_GEO(dgg, wff_sum$cell)$lat_deg 
wff_sum$cell_lon <- dgSEQNUM_to_GEO(dgg, wff_sum$cell)$lon_deg 

# Get the grid cell boundaries 
eft_grid <- dggridR::dgcellstogrid(dgg, wff_sum$cell) 

#Update the grid cells' properties to include sc values
eft_grid <- merge(eft_grid, wff_sum, by.x="seqnum", by.y="cell")
```


Identify which cells fall within and outside of the eastern redcedar
range extent (ERC_range_sf).

-   Convert the wff dataframe into an sf object.
-   Geometry for sc_sf represents the center point of the icosahedron
    cell within which the observation took place.
-   Geometry for ERC_range_sf represents the range boundaries for
    eastern red cedar tree

```{r}
#convert wff_sum grid data into an sf object 
wff_sum_sf <- st_as_sf(eft_grid, 
                     coords = c("cell_lon", "cell_lat"), 
                     crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

# apply st_intersects to wff data
wff_sf_ERC = lengths(st_intersects(wff_sum_sf, ERC_range_sf)) > 0

# use sc_sf_ERC to filter the sc data to only include observations WITHIN ERC RANGE extent
wff_ERC_in = wff_sum_sf %>%
  filter(geometry %in% geometry[wff_sf_ERC]) %>%
  mutate(ERC="within_range")
```

Outlier frequency values causes the map to look a bit strange. Label them here so we can plot differently
```{r}
#define outlier value for frequency using the percentiles method
lower_bound <- quantile(wff_ERC_in$freq, 0.025)
upper_bound <- quantile(wff_ERC_in$freq, 0.975)

# subset the dataframe to rows where freq is an upper outlier
# the upper outliers are the ones that will mess with the visual map fill
wff_ERC_outliers <- wff_ERC_in %>%
  filter(freq > upper_bound)
```



I want to see a map showing where the winter fruit feeder species are located, during the fruiting season (November-February). Plot where fill is count/effort, so it shows some measure of the likelihood that bird is present or abundant, given the total eBirder effort in the cell. Only plot occurrence for the cells that fall within the J. virginiana range. 

Let's plot summed winter fruit feeder frequency on a map. 
Note: Here we use the wff_ERC_in to show frequency as fill values. 
We overlay a polygon showing the ERC range extent on top of the hexagons.

## Winter fruit feeder map
```{r}
#  frequency of winter fruit feeders
ggplot() + 
  geom_sf(data = wff_ERC_in %>% filter(diet == "winter fruit feeder" &
                                         freq < upper_bound),
          aes(fill=freq), color = alpha("white", 0.4 )) +
  geom_sf(data=wff_ERC_outliers %>% filter(diet == "winter fruit feeder"), 
          col="red", fill="lemonchiffon") +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black", linewidth=0.10) +
  geom_sf(data = ERC_range_sf, fill = NA, color ="black", linewidth=0.25) +  
  scale_fill_viridis_c(option = "plasma", limits=c(0,upper_bound)) +
  coord_sf(xlim = c(-106, -70), ylim = c(29, 49)) + #set limits without distorting boundaries
  labs(title = "Winter fruit feeder frequency (Oct-Mar)") +
  theme_void() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          legend.position="bottom") +
  facet_wrap(~species, ncol=3)

ggsave(filename = here("figs/winter-fruit-feeder_map.png"), height = 11, width=8, bg="white")
```

## Fruit feeder map
```{r}
#  frequency of fruit feeders
ggplot() + 
  geom_sf(data = wff_ERC_in %>% filter(diet == "fruit feeder" & 
                                         freq < upper_bound), 
          aes(fill=freq), color = alpha("white", 0.4 )) +
    geom_sf(data=wff_ERC_outliers %>% filter(diet == "fruit feeder"), 
          col="red", fill="lemonchiffon") +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black", linewidth=0.10) +
  geom_sf(data = ERC_range_sf, fill = NA, color ="black", linewidth=0.25) +  
  scale_fill_viridis_c(option = "plasma", limits=c(0,upper_bound) ) +
  coord_sf(xlim = c(-106, -70), ylim = c(29, 49)) + #set limits without distorting boundaries
  labs(title = "fruit feeder frequency (Oct-Mar)") +
  theme_void() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          legend.position="bottom") +
  facet_wrap(~species, ncol=3)

ggsave(filename = here("figs/fruit-feeder_map.png"), height = 11, width=8, bg="white")
```

## opportunistic feeder map
```{r}
#  frequency of fruit feeders
ggplot() + 
  geom_sf(data = wff_ERC_in %>% filter(diet == "opportunistic" &
                                         freq < upper_bound), 
          aes(fill=freq), color = alpha("white", 0.4 )) +
    geom_sf(data=wff_ERC_outliers %>% filter(diet == "opportunistic"), 
          col="red", fill="lemonchiffon") +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black", linewidth=0.10) +
  geom_sf(data = ERC_range_sf, fill = NA, color ="black", linewidth=0.25) +  
  scale_fill_viridis_c(option = "plasma", limits=c(0,upper_bound)) +
  coord_sf(xlim = c(-106, -70), ylim = c(29, 49)) + #set limits without distorting boundaries
  labs(title = "opportunistic feeder frequency (Oct-Mar)") +
  theme_void() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          legend.position="bottom") +
  facet_wrap(~species, ncol=3)

ggsave(filename = here("figs/opportunistic-feeder_map.png"), height = 11, width=8, ncol=3, bg="white")
```

