---
title: "Occurrence Range Maps"
author: "Maximilian Wisnefski"
date: '2023-06-05'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rgdal)
library(sf)
library(s2)
library(geosphere)
library(rnaturalearth)
#devtools::install_github("ropensci/rnaturalearthhires")
library(rnaturalearthhires)
```


This is just for the purpose of calculating migration distances for the avian traits table. However, I also made some cool maps of the occurrence ranges.
```{r}
#commenting these out since we can just use .rds with the info that we need
#and because github repo doesn't actually have .gdb file on it (too big to push)
#bird_occ_map <- st_read("BOTW.gdb", layer = "All_Species")

#layers <- st_layers(dsn = "BOTW.gdb")
```


```{r}
#commenting this out for the same reason

#filt_birds <- bird_occ_map[bird_occ_map$sci_name %in% c("Bombycilla cedrorum", "Turdus migratorius", "Hylocichla mustelina", "Setophaga coronata", "Cyanocitta cristata", "Sturnus vulgaris", "Sialia sialis", "Mimus polyglottos", "Dryobates pubescens", "Sturnella magna", "Sitta carolinensis", "Haemorhous purpureus", "Cardinalis cardinalis", "Junco hyemalis", "Corvus brachyrhynchos"), ]
```



```{r}
#saveRDS(filt_birds, "occurrence_map.rds")
birds <- readRDS("data_2023/bird_occurrence_range_maps/occurrence_map.rds")
```

```{r}
#row vals of 2 and 3 for "seasonal" col represent polygons for breeding season range and non-breeding season range respectively, so we filter for those two vals
filt_birds <- birds[birds$seasonal %in% c(2, 3), ]
```


```{r}
#sf_use_s2(TRUE)
#calculating the centroid breeding/non-breeding ranges for every bird in df 
filt_birds$centroid <- st_centroid(st_make_valid(filt_birds$Shape), of_largest_polygon = FALSE)
```


```{r}
#was initially using this shapefile but the below chunk provides an easier way of getting a background map of north America
#US_CAN_spdf <- readOGR(dsn = "../data/USA_Canada_ShapefileMerge", layer = "USA_Canada_ShapefileMerge", verbose = TRUE)
```


```{r}
#getting maps of North America and zooming in to desired lat/lon ranges
state_prov <- rnaturalearth::ne_states(c("united states of america", "canada", "mexico", "honduras", "el salvador", "guatemala", "costa rica"), returnclass = "sf")
state_prov <- st_crop(state_prov, xmin = -170, xmax = -50, ymin = 5, ymax = 70)
```

## Cedar Waxwing

```{r}
#plotting breeding and non-breeding ranges of cedar waxwing
#use col2rgb("color") to find right numbers for custom color you want
mycol1 <- rgb(173, 216, 230, max = 255, alpha = 170, names = "color")
mycol2 <- rgb(255, 165, 162, max = 255, alpha = 170, names = "color")

#TO DO: should probably find a shapefile that includes Mexico
par(mar = c(0,0,2.5,0))

plot(state_prov$geometry,
     main = "Map of US and Canada\nWith Cedar Waxwing Ranges",
     col = "white", lwd = 0.25, xlim = c(-130,-50))
plot(filt_birds[1:1, ]$Shape, col = mycol1, add = TRUE) 
plot(filt_birds[2:2, ]$Shape, col = mycol2, add = TRUE) 

#plotting the centroids of the range
plot(filt_birds[1:2, ]$centroid, col = "darkgreen", pch = 16, add = TRUE)
```


```{r}
#distance between centroids of ranges for Cedar Waxwing (in meters)
distHaversine(c(filt_birds$centroid[[1]][1], filt_birds$centroid[[1]][2]), c(filt_birds$centroid[[2]][1], filt_birds$centroid[[2]][2]))
```


## Eastern Bludebird

```{r}
par(mar = c(0,0,2.5,0))
plot(state_prov$geometry,
     main = "Map of US and Canada\nWith Eastern Bluebird Ranges",
     col = "white", lwd = 0.25)

plot(filt_birds[3:3, ]$Shape, col = mycol1, add = TRUE) 
plot(filt_birds[4:4, ]$Shape, col = mycol2, add = TRUE) 

#plotting the centroids of the range
plot(filt_birds[3:4, ]$centroid, col = "darkgreen", pch = 16, add = TRUE)
```

```{r}
#distance between centroids of ranges for Cedar Waxwing (in meters)
distHaversine(c(filt_birds$centroid[[3]][1], filt_birds$centroid[[3]][2]), c(filt_birds$centroid[[4]][1], filt_birds$centroid[[4]][2]))
```


## American Robin

```{r}
par(mar = c(0,0,2.5,0))
plot(state_prov$geometry, 
     main = "Map of US and Canada\nWith American Robin Ranges",
     col = "white", lwd = 0.25)

plot(filt_birds[6:6, ]$Shape, col = mycol1, add = TRUE) 
plot(filt_birds[5:5, ]$Shape, col = mycol2, add = TRUE) 

#plotting the centroids of the range
plot(filt_birds[5:6, ]$centroid, col = "darkgreen", pch = 16, add = TRUE)
```


```{r}
#distance between centroids of ranges for American Robin (in meters)
distHaversine(c(filt_birds$centroid[[5]][1], filt_birds$centroid[[5]][2]), c(filt_birds$centroid[[6]][1], filt_birds$centroid[[6]][2]))
```


## Wood Thrush

```{r}
#easier to just make subset
wood_thrush <- filt_birds[filt_birds$sci_name %in% c("Hylocichla mustelina"), ]


par(mar = c(0,0,2.5,0))
plot(state_prov$geometry,
     main = "Map of US and Canada\nWith Wood Thrush Ranges",
     col = "white", lwd = 0.25)

plot(wood_thrush[2:2, ]$Shape, col = mycol1, add = TRUE) 
plot(wood_thrush[1:1, ]$Shape, col = mycol2, add = TRUE) 

#plotting the centroids of the range
plot(wood_thrush[1:2, ]$centroid, col = "darkgreen", pch = 16, add = TRUE)
```


```{r}
#distance between centroids of ranges for Wood Thrush (in meters)
distHaversine(c(wood_thrush$centroid[[1]][1], wood_thrush$centroid[[1]][2]), c(wood_thrush$centroid[[2]][1], wood_thrush$centroid[[2]][2]))
```


## Yellow-Rumped Warbler

```{r}
#easier to just make subset
YR_warbler <- filt_birds[filt_birds$sci_name %in% c("Setophaga coronata"), ]


par(mar = c(0,0,2.5,0))
plot(state_prov$geometry,
     main = "Map of US and Canada\nWith Yellow-Rumped Warbler Ranges",
     col = "white", lwd = 0.25)

plot(YR_warbler[1:1, ]$Shape, col = mycol1, add = TRUE) 
plot(YR_warbler[2:2, ]$Shape, col = mycol2, add = TRUE) 

#plotting the centroids of the range
plot(YR_warbler[1:2, ]$centroid, col = "darkgreen", pch = 16, add = TRUE)
```


```{r}
#distance between centroids of ranges (in meters)
distHaversine(c(YR_warbler$centroid[[1]][1], YR_warbler$centroid[[1]][2]), c(YR_warbler$centroid[[2]][1], YR_warbler$centroid[[2]][2]))
```


## Blue Jay

```{r}
#easier to just make subset
blue_jay <- filt_birds[filt_birds$sci_name %in% c("Cyanocitta cristata"), ]


par(mar = c(0,0,2.5,0))
plot(state_prov$geometry,
     main = "Map of US and Canada\nWith Blue Jay Ranges",
     col = "white", lwd = 0.25, border = 1)

plot(blue_jay[2:2, ]$Shape, col = mycol1, add = TRUE) 
plot(blue_jay[1:1, ]$Shape, col = mycol2, add = TRUE) 

#plotting the centroids of the range
plot(blue_jay[1:2, ]$centroid, col = "darkgreen", pch = 16, add = TRUE)
```


```{r}
#distance between centroids of ranges (in meters)
distHaversine(c(blue_jay$centroid[[1]][1], blue_jay$centroid[[1]][2]), c(blue_jay$centroid[[2]][1], blue_jay$centroid[[2]][2]))
```


## European Starling

```{r}
#easier to just make subset
euro_star <- filt_birds[filt_birds$sci_name %in% c("Sturnus vulgaris"), ]


par(mar = c(0,0,2.5,0))
plot(state_prov$geometry,
     main = "Map of US and Canada\nWith European Starlings Ranges",
     col = "white", lwd = 0.25)

plot(euro_star[4:4, ]$Shape, col = mycol1, add = TRUE) 
plot(euro_star[3:3, ]$Shape, col = mycol2, add = TRUE) 

#plotting the centroids of the range
plot(euro_star[3:4, ]$centroid, col = "darkgreen", pch = 16, add = TRUE)
```


```{r}
#distance between centroids of ranges (in meters)
distHaversine(c(euro_star$centroid[[3]][1], euro_star$centroid[[3]][2]), c(euro_star$centroid[[4]][1], euro_star$centroid[[4]][2]))
```


## Eastern Meadowlark 
Unfortunately, the dataset only features the Eastern Meadowlark's breeding range

```{r}
#easier to just make subset
meadow <- filt_birds[filt_birds$sci_name %in% c("Sturnella magna"), ]


par(mar = c(0,0,2.5,0))
plot(state_prov$geometry,
     main = "Map of US and Canada\nWith Eastern Meadowlark Ranges",
     col = "white", lwd = 0.25, border = 1)

plot(meadow[1:1, ]$Shape, col = mycol1, add = TRUE) 
plot(meadow[2:2, ]$Shape, col = mycol2, add = TRUE) 

#plotting the centroids of the range
plot(meadow[1:2, ]$centroid, col = "darkgreen", pch = 16, add = TRUE)
```

## Purple Finch

```{r}
#easier to just make subset
finch <- filt_birds[filt_birds$sci_name %in% c("Haemorhous purpureus"), ]


par(mar = c(0,0,2.5,0))
plot(state_prov$geometry,
     main = "Map of US and Canada\nWith Blue Jay Ranges",
     col = "white", lwd = 0.25)

plot(finch[2:2, ]$Shape, col = mycol1, add = TRUE) 
plot(finch[1:1, ]$Shape, col = mycol2, add = TRUE) 

#plotting the centroids of the range
plot(finch[1:2, ]$centroid, col = "darkgreen", pch = 16, add = TRUE)
```


```{r}
#distance between centroids of ranges (in meters)
distHaversine(c(finch$centroid[[1]][1], finch$centroid[[1]][2]), c(finch$centroid[[2]][1], finch$centroid[[2]][2]))
```


## Dark-Eyed Junco

```{r}
#easier to just make subset
junco <- filt_birds[filt_birds$sci_name %in% c("Junco hyemalis"), ]


par(mar = c(0,0,2.5,0))
plot(state_prov$geometry,
     main = "Map of US and Canada\nWith Dark-Eyed Junco Ranges",
     col = "white", lwd = 0.25)

plot(junco[2:2, ]$Shape, col = mycol1, add = TRUE) 
plot(junco[1:1, ]$Shape, col = mycol2, add = TRUE) 

#plotting the centroids of the range
plot(junco[1:2, ]$centroid, col = "darkgreen", pch = 16, add = TRUE)
```


```{r}
#distance between centroids of ranges (in meters)
distHaversine(c(junco$centroid[[1]][1], junco$centroid[[1]][2]), c(junco$centroid[[2]][1], junco$centroid[[2]][2]))
```


## American Crow

```{r}
#easier to just make subset
crow <- filt_birds[filt_birds$sci_name %in% c("Corvus brachyrhynchos"), ]


par(mar = c(0,0,2.5,0))
plot(state_prov$geometry,
     main = "Map of US and Canada\nWith American Crow Ranges",
     col = "white", lwd = 0.25)

plot(crow[2:2, ]$Shape, col = mycol1, add = TRUE) 
plot(crow[1:1, ]$Shape, col = mycol2, add = TRUE) 

#plotting the centroids of the range
plot(crow[1:2, ]$centroid, col = "darkgreen", pch = 16, add = TRUE)
```


```{r}
#distance between centroids of ranges (in meters)
distHaversine(c(crow$centroid[[1]][1], crow$centroid[[1]][2]), c(crow$centroid[[2]][1], crow$centroid[[2]][2]))
```



