---
title: "process-eBird-data"
author: "Sarah Supp"
date: "2/19/2021"
output: html_document
---

##Code for cedar waxwing project, modified from eBird migration project (Supp et al. 2015)
(c) 2020-2025, Niu and Supp
supps@denison.edu
Denison University
Code is under development, part of NSF Multi-Institution Collaborative Award (2019-25)

Birds evaluated include: 
* Cedar Waxwing (CEWA) _Bombycilla cedrorum_
* Robin (ROBI) _Turdus migratorius_
* Wood Thrush (WOTH) _Hylocichla mustelina_
* Yellow-rumped Warbler (YEWA) _setophaga coronata_
* Blue Jay (BLJA) _Cyanocitta cristata_
* European Starling (EUST) _Sturnus vulgaris_
* Eastern Bluebird (EABL) _Sialia sialis_
* Northern Mockingbird (NOMO) _Mimus polyglottos_
* Downy Woodpecker (DOWO) _Dryobates pubescens_
* Eastern Meadowlark (EAME) _Sturnella magna_
* White-breasted Nuthatch (WHNU) _Sitta carolinensis_
* Purple Finch (PUFI) _Haemorhous purpureus_
* Northern Cardinal (NOCA) _Cardinalis cardinalis_
* Dark-eyed Junco (DAJU) _Junco hyemalis_
* American Crow (AMCR) _Corvus brachyrhynchos_


```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
```

If starting from scratch, input the three raw data files and merge together.
You need to bring in: 
1. a file with the bird observations (ebrd2.c)
2. a file with the total eBirder "effort" (dat; effort-count.RData)
3. a file with all the center locations for the POLYFID grid cells (locs)

*Load datasets* 
Load the eBird effort and location files, from FAL (January 2025)
Data files should contain values from the western hemisphere
```{r}
# raw eBird data, queried by FAL
  # will load as dat, FAL named previously
load(here("data/1-raw_eBird_data/effort-count.RData"))

# coordinates for the icosahedron cells
 # will load as locs, FAL named previously
load(here("data/1-raw_eBird_data/locs.RData")) 
```


*For all 15 species .RData files*
This code chunk will identify all 15 species files, bring them in one by one, and loop through the merge on polygon ID (POLYFID) with the total eBirder effort, add the center lat-long coordinates for each hexagonal icosahedron cell, and write to .RDS file for use in the next steps.
```{r}
spfiles = list.files(path=here("data/1-raw_eBird_data/species/"), pattern=glob2rx("*.RData"))

for (i in spfiles) {
  # These RData files have one object and it is named "dat2" 
  load(paste0(here("data/1-raw_eBird_data/species/"),(i)))
  
  # use merge to annotate the species files with the effort values
  datmerged <- merge(dat2, dat)
  
  # use merge to annotate the species counts and effort with the polygon coordinates (lat, long)
  datmerged <- merge(datmerged, locs)

  # grab species name from the filename and add it as a column
  spname <- sub("\\.[[:alnum:]]+$", "", i)
  datmerged$species <- spname
  
  # save file, also adding the species name, into a new folder with the species + effort merged datasets
  saveRDS(datmerged, file = here(paste0("data/2-intermediate_eBird_data/species_effort/dat_effort_", spname, ".rds")))
  
  print(paste0(spname, " has ", nrow(datmerged), " rows and is complete."))
}
```
